Generating Random Logic Programs Using Constraint Programming

### 2020-01-19 paper/paper.tex
Fix the appearance of algorithms

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
%\usepackage[UKenglish]{babel}
%\usepackage[UKenglish]{isodate}
%\usepackage[utf8]{inputenc}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}

\usetikzlibrary{arrows.meta}

\newtheorem{constraint}{Constraint}

@@

\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\arrayd}[3]{\variable{{#1}[}{#2}\variable{]} \in {#3}}
\newcommand{\arrayt}[3]{\variable{{#3}[}{#2}\variable{] {#1}}}
\newcommand{\logical}[1]{{\normalfont \texttt{#1}}}

@@

\DeclareMathOperator{\Determined}{Determined}
\DeclareMathOperator{\Undetermined}{Undetermined}
\DeclareMathOperator{\AlmostDetermined}{AlmostDetermined}

%\relpenalty=10000
%\binoppenalty=10000
\def\multiset#1#2{\ensuremath{\left(\kern-.3em\left(\genfrac{}{}{0pt}{}{#1}{#2}\right)\kern-.3em\right)}}

\begin{document}

\title{Generating Random Logic Programs Using Constraint Programming}
\author{Paulius Dilkas\orcidID{0000-1111-2222-3333}}
\authorrunning{P. Dilkas}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.

\institute{University of Edinburgh, Edinburgh, United Kingdom\\
  \email{p.dilkas@sms.ed.ac.uk}}
\maketitle

\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{Constraint Programming \and Logic Programming \and Probabilistic Logic
  Programming.}
\end{abstract}

\section{Introduction}

@@

We also use Choco~4.10.2 \cite{choco}. This works with both Prolog
\cite{DBLP:books/daglib/0041598} and ProbLog \cite{DBLP:conf/ijcai/RaedtKT07}.
%Tested with SWI-Prolog \cite{DBLP:journals/tplp/WielemakerSTL12}.

\subsection{Parameters}

Parameters:
\begin{itemize}
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$ (including zero, but assuming that at
  least one predicate has non-zero arity),
  \begin{itemize}
  \item maximum arity $\maxArity{} \coloneqq \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
  \begin{itemize}
  \item Each of them can be empty, but $|\constants{}| + |\variables{}| > 0$.
  \end{itemize}
\item a list of probabilities that are randomly assigned to clauses,
\item option to forbid all cycles or just negative cycles,
\item $\maxNumNodes{} \ge 1$: maximum number of nodes in the tree representation
  of a clause,
\item $\maxNumClauses{} \ge |\predicates{}|$: maximum number of clauses in a
  program,
\item maximum number of solutions,
\end{itemize}

We also define $\tokens{} = \{ \neg, \land, \lor, \top \}$. All decision
variables of the model are contained in two arrays:
\begin{itemize}
\item $\arrayt{bodiesOfClauses}{\maxNumClauses{}}{Body}$,
\item $\arrayt{headsOfClauses}{\maxNumClauses{}}{Head}$
\end{itemize}

@@

\section{Heads of Clauses}

\begin{definition}
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item a $\variable{predicate} \in \predicates \cup \{ \Box \}$.
  \item and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}}
    \cup \{ \Box \}$
  \end{itemize}
\end{definition}
The reason why $\Box$ must be a separate value will become clear in
\cref{sec:variable_symmetry}.

\begin{definition} \label{def:arity}
  The \variable{predicate}'s $\variable{arity} \in [0, \maxArity{}]$ can then be
  defined using the \variable{table} constraint as
  \[
    \variable{arity} = \begin{cases}
      \text{the arity of } \variable{predicate} & \text{if }
      \variable{predicate} \in \predicates{}\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

@@

\begin{constraint}
  Each predicate gets at least one clause. Let
  \[
    P = \{ h.\variable{predicate} \mid h \in \variable{headsOfClauses} \}.
  \]
  Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      \variable{numPredicates} + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      \variable{numPredicates} & \text{otherwise.}
    \end{cases}
  \]
  Here, $\variable{nValues}(P)$ counts the number of unique values in $P$.
\end{constraint}

\section{Bodies of Clauses}

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item $\arrayd{treeStructure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ such
    that:
    \begin{itemize}
    \item $\variable{treeStructure}[i] = i$: the $i$-th node is a root.
    \item $\variable{treeStructure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item $\arrayt{treeValues}{\maxNumNodes{}}{Node}$.
  \end{itemize}
\end{definition}

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (-0.375, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1.5, -1) {$\neg$};
    \node[draw,circle,gray,text=black] (and) at (0.75, -1) {$\land$};
    \node[draw,gray,text=black] (P) at (-1.5, -2) {$\mathsf{P}(X)$};
    \node[draw,gray,text=black] (Q) at (0, -2) {$\mathsf{Q}(X)$};
    \node[draw,gray,text=black] (R) at (1.5, -2) {$\mathsf{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \caption{A tree representation of the formula from \cref{example:formula}}
  \label{fig:example_tree}
\end{figure}

\begin{example} \label{example:formula}
  Let $\maxNumNodes{} = 8$. Then $\neg\mathsf{P}(X) \lor (\mathsf{Q}(X)
  \land \mathsf{P}(X))$ corresponds to the tree in \cref{fig:example_tree} and
  can be encoded as:
  \begin{alignat*}{8}
    \variable{treeStructure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
    \variable{treeValues} &= [{\lor}, &&{\neg}, &&{\land}, \mathsf{P}(&&X), \mathsf{Q}(&&X), \mathsf{P}(&&X), &&\top, &&\top],\\
    \variable{numNodes} &= 6, && && && && && && &&\\
    \variable{numTrees} &= 3. && && && && && && &&
  \end{alignat*}
  In the rest of this section, we will describe how the elements of
  $\variable{treeValues}$ are encoded and list a series of constraints that make
  this representation unique.
\end{example}

\subsection{Nodes}

@@

\subsection{Constraints}

\begin{definition}
  We define $\variable{numTrees} \in \{ 1, \dots, \maxNumNodes{} \}$ to count
  the number of trees in our representation of a clause using the
  $\variable{tree}(\variable{treeStructure}, \variable{numTrees})$\footnote{This
    constraint uses dominator-based filtering by Fages and Lorca
    \cite{DBLP:conf/cp/FagesL11}.} constraint.
\end{definition}

@@

\begin{constraint}
  $\variable{treeStructure}[0] = 0$.
\end{constraint}

\begin{constraint}
  \variable{treeStructure} is sorted.
\end{constraint}

\begin{constraint}
  For $i = 1, \dots, \maxNumNodes{} - 1$, if $i \ge \variable{numNodes}$, then
  \[
    \variable{treeStructure}[i] = i \quad \text{and} \quad
    \variable{treeValues}[i].\variable{name} = \top,
  \]
  else
  \[
    \variable{treeStructure}[i] < i.
  \]
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \begin{align*}
    \variable{count}(i, \variable{treeStructure}_{-i}) = 0 &\iff \variable{treeValues}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    \variable{count}(i, \variable{treeStructure}_{-i}) = 1 &\iff \variable{treeValues}[i].\variable{name} = \neg,\\
    \variable{count}(i, \variable{treeStructure}_{-i}) > 1 &\iff \variable{treeValues}[i].\variable{name} \in \{ \land, \lor \},
  \end{align*}
  where $\variable{treeStructure}_{-i}$ denotes array $\variable{treeStructure}$
  with position $i$ skipped.
\end{constraint}
Each constraint corresponds to node $i$ having no children, one child, and
multiple children, respectively.

\begin{constraint}
  For $i = 0, \dots, \maxNumNodes{} - 1$,
  \[
    \variable{treeStructure}[i] \ne i \implies
    \variable{treeValues}[i].\variable{name} \ne \top.
  \]
\end{constraint}

\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{headsOfClauses}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodiesOfClauses}[i].\variable{numNodes} = 1,
  \]
  and
  \[
    \variable{bodiesOfClauses}[i].\variable{treeValues}[0].\variable{name} =
    \top.
  \]
\end{constraint}

\section{Eliminating Variable Symmetries} \label{sec:variable_symmetry}

@@

  Then we can use the $\variable{setsIntsChanneling}$ constraint
  to define $\variable{occurrences}[|\constants{}| + |\variables{}| + 1]$ as an
  array of subsets of $\{ 0, \dots, N-1 \}$ such that for all $i = 0, \dots, N
  - 1$, and $t \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    i \in \variable{occurrences}[t] \quad \iff \quad
    \variable{terms}[i] = t
  \]
\end{definition}

\begin{definition}
  We define $\arrayd{introductions}{|\variables{}|}{\{ 0, \dots, N \}}$ such
  that for $v \in \variables{}$,
  \[
    \variable{introductions}[v] = \begin{cases}
      1 + \min \variable{occurrences}[v] & \text{if }
      \variable{occurrences}[v] \ne \emptyset\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}

\begin{constraint}
  $\variable{introductions}$ are sorted.
\end{constraint}

\begin{example}
  Let $\constants{} = \emptyset$, $\variables{} = \{ X, Y, Z \}$, $\maxArity{} =
  2$, $\maxNumNodes{} = 3$, and consider the clause
  \[
    \mathsf{sibling}(X, Y) \gets \mathsf{parent}(X, Z) \land
    \mathsf{parent}(Y, Z).
  \]
  Then $\variable{terms} = [X, Y, \Box, \Box, X, Z, Y, Z]$ (the boxes represent
  the conjunction node), $\variable{occurrences} = [\{ 0, 4 \}, \{ 1, 6 \},
  \{ 5, 7 \}, \{ 2, 3 \}]$, and $\variable{introductions} = [0, 1, 5]$.
\end{example}

\subsection{Redundant Constraints}

@@

\begin{constraint}
  For $u \ne v \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    \variable{occurrences}[u] \cap \variable{occurrences}[v] = \emptyset.
  \]
\end{constraint}

\begin{constraint}
  $\variable{allDifferentExcept0}(\variable{introductions})$.
\end{constraint}

\begin{constraint}
  For $v \in \variables{}$,
  \[
    \variable{introductions}[v] \ne 0 \quad \implies \quad
    \variable{introductions}[v] - 1 \in \variable{occurrences}[v].
  \]
\end{constraint}

\begin{definition}
  We define an auxiliary set variable
  \[
    \variable{potentialIntroductions} \subseteq \{ 0, \dots, N \}
  \]
  by: for $i = 0, \dots, \maxNumNodes{} - 1$, if
  $\variable{treeValues}[i].\variable{name} \not\in \predicates{}$, then
  \[
    \variable{potentialIntroductions} \cap \{ \maxArity{} \times (i + 1) + j + 1
    \mid j = 0, \dots, \maxArity{} - 1 \} = \emptyset.
  \]
\end{definition} % TODO: explain these numbers
In other words, if the node is not a predicate, a variable cannot be introduced
as one of its `arguments'.

\begin{constraint}
  For $v \in \variables{}$, $\variable{introductions}[v] \in
  \variable{potentialIntroductions}$.
\end{constraint}

\section{Counting Programs}

@@

We will first consider clauses with gaps, i.e., without taking variables and
constants into account. Let $T(n, a)$ denote the number of possible clause
bodies with $n$ nodes and total arity $a$. Then $T(1, a)$ is the number of
predicates in $\predicates{}$ with arity $a$, and the following recursive
definition can be applied for $n > 1$:
\[
  T(n, a) = T(n-1, a) + 2\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
      c_i \ge 1 \text{ for all } i}} \sum_{\substack{d_1 + \dots + d_k = a,\\
    d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\]
The first term here represents negation, i.e., negating an expression consumes
one node but otherwise leaves the task unchanged. If the first operation is not
negation, then it must be either conjunction or disjunction (hence the
coefficient `2'). In the first sum, $k$ represents the number of children of the
root node, and each $c_i$ is the number of nodes dedicated to child $i$. Thus,
the first sum iterates over all possible ways to partition the remaining $n-1$
nodes. Similarly, the second sum considers every possible way to partition the
total arity $a$ across the $k$ children nodes.

@@

The number of ways to fill $n$ gaps with terms can be expressed as
\[
  P(n) = |\constants{}|^n + \sum_{\substack{1 \le k \le |\variables{}|, \\ 0 =
      s_0 < s_1 < \dots < s_k < s_{k+1} = n+1}} \prod_{i=0}^k (|\constants{}| +
  i)^{s_{i+1} - s_i - 1}.
\]
The first term is simply the number of ways to fill all $n$ gaps with
constants. The parameter $k$ is the number of variables used in the clause, and
$s_1, \dots, s_k$ mark the first occurrence of each variable. For each gap
between any two introductions (or before the first introduction, or after the last
introduction), we have $s_{i+1}-s_i-1$ spaces to be filled with any of the
$\constants{}$ constants or any of the $i$ already-introduced variables.

@@

\section{Predicate Independence}

@@

\begin{figure}[t]
  \centering
  \begin{minipage}{.5\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
      \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
      \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
      \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
      \draw[gray] (and) -- (parent);
      \draw[gray] (and) -- (not);
      \draw[gray] (not) -- (mother);
    \end{tikzpicture}
    \caption{The tree representation of the body of Clause~\eqref{eq:example_clause}.}
    \label{fig:example_tree2}
  \end{minipage}
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
      \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
    \end{tikzpicture}
    \caption{The predicate dependency graph of the program in \cref{ex:program}.
      Positive edges are labeled with `$+$', and negative edges with `$-$'.}
    \label{fig:predicate_dependencies}
  \end{minipage}
\end{figure}

@@

  \For{$(d_1, d_2) \in \getDependencies{$p_1$, \logical{false}} \times
      \getDependencies{$p_2$, \logical{false}}$ such that $d_1.\predicate =
      d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined$ {\bf and} $d_2$ {\bf is}

>>>>

\documentclass[letterpaper]{article}
\usepackage{uai2019}
\usepackage[margin=1in]{geometry}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage{siunitx}
\usepackage{amsthm}
\usepackage[round]{natbib}

\newtheorem{constraint}{Constraint}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

@@

\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother

\newcommand{\logical}[1]{{\normalfont \texttt{#1}}}
\newcommand{\variable}[1]{\texttt{\textup{#1}}}
\newcommand{\arrayd}[3]{\variable{{#1}[}{#2}\variable{]} \in {#3}}
\newcommand{\arrayt}[3]{\variable{{#3}[}{#2}\variable{] {#1}}}

@@

\DeclareMathOperator{\Determined}{\Delta}
\DeclareMathOperator{\Undetermined}{\Upsilon}
\DeclareMathOperator{\AlmostDetermined}{\Gamma}

\usetikzlibrary{arrows.meta}
\Crefname{constraint}{Constraint}{Constraints}
\relpenalty=10000
\binoppenalty=10000
\def\multiset#1#2{\ensuremath{\left(\kern-.3em\left(\genfrac{}{}{0pt}{}{#1}{#2}\right)\kern-.3em\right)}}

% Set the typeface to Times Roman
\usepackage{times}

\title{Generating Random Logic Programs Using Constraint Programming}

\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
          % UAI  reviewing is double-blind.

% The author names and affiliations should appear only in the accepted paper.
%
%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
%alternate address.}} \\
%Computer Science Dept. \\
%Cranberry University\\
%Pittsburgh, PA 15213 \\
%\And
%{\bf Coauthor}  \\
%Affiliation          \\
%Address \\
%\And
%{\bf Coauthor}   \\
%Affiliation \\
%Address    \\
%(if needed)\\
%}

% \begin{figure}[h]
% \vspace{1in}
% \caption{Sample Figure Caption}
% \end{figure}

% \begin{table}[h]
% \caption{Sample Table Title}
% \label{sample-table}
% \begin{center}
% \begin{tabular}{ll}
% \multicolumn{1}{c}{\bf PART}  &\multicolumn{1}{c}{\bf DESCRIPTION} \\
% \hline \\
% Dendrite         &Input terminal \\
% Axon             &Output terminal \\
% Soma             &Cell body (contains cell nucleus) \\
% \end{tabular}
% \end{center}
% \end{table}

\begin{document}
\bibliographystyle{plainnat}
\maketitle

\begin{abstract}
The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
both left and right-hand margins.  Use 10~point type, with a vertical
spacing of 11~points.  {\bf Abstract} must be centered, bold, and in
point size 12. Two line spaces precede the Abstract. The Abstract must
be limited to one paragraph.
\end{abstract}

\section{INTRODUCTION}

@@

We also use Choco~4.10.2 \citep{choco}. This works with both Prolog
\citep{DBLP:books/daglib/0041598} and ProbLog \citep{DBLP:conf/ijcai/RaedtKT07}.
%Tested with SWI-Prolog \citep{DBLP:journals/tplp/WielemakerSTL12}.

\subsection{PARAMETERS}

Parameters:
\begin{itemize}
\item a list of predicates $\predicates{}$,
\item a list of their arities $\arities{}$ (including zero, but assuming that at
  least one predicate has non-zero arity),
  \begin{itemize}
  \item \emph{maximum arity} $\maxArity{} \coloneqq \max \arities{}$.
  \end{itemize}
\item a list of variables $\variables{}$,
\item and a list of constants $\constants{}$.
  \begin{itemize}
  \item Each of them can be empty, but $|\constants{}| + |\variables{}| > 0$.
  \end{itemize}
\item a list of probabilities that are randomly assigned to clauses,
\item option to forbid all cycles or just negative cycles,
\item $\maxNumNodes{} \ge 1$: maximum number of nodes in the tree representation
  of a clause,
\item $\maxNumClauses{} \ge |\predicates{}|$: maximum number of clauses in a
  program,
\end{itemize}

We also define $\tokens{} = \{ \neg, \land, \lor, \top \}$. All decision
variables of the model are contained in two arrays:
$\arrayt{bodies}{\maxNumClauses{}}{Body}$ and
$\arrayt{heads}{\maxNumClauses{}}{Head}$.

@@

A random selection of probabilities is added at the end in case a
probabilistic program is desired. Regular logic programs can be generated by
setting the list of probabilities to $\{ 1 \}$.

\section{HEADS OF CLAUSES}

\begin{definition}
  The \emph{head} of a clause is composed of
  \begin{itemize}
  \item a $\variable{predicate} \in \predicates \cup \{ \Box \}$.
  \item and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}}
    \cup \{ \Box \}$
  \end{itemize}
\end{definition}
The reason why $\Box$ must be a separate value will become clear in
\cref{sec:variable_symmetry}.

\begin{definition} \label{def:arity}
  The \variable{predicate}'s $\variable{arity} \in [0, \maxArity{}]$ can then be
  defined using the \variable{table} constraint as the arity of the
  $\variable{predicate}$ if $\variable{predicate} \in \predicates{}$, and zero
  otherwise.
\end{definition}

@@

\begin{constraint}
  Each predicate gets at least one clause. Let
  \[
    P = \{ h.\variable{predicate} \mid h \in \variable{heads} \}.
  \]
  Then
  \[
    \variable{nValues}(P) =
    \begin{cases}
      |\predicates{}| + 1 & \text{if } \variable{count}(\Box, P) > 0 \\
      |\predicates{}| & \text{otherwise.}
    \end{cases}
  \]
  Here, $\variable{nValues}(P)$ counts the number of unique values in $P$.
\end{constraint}

\section{BODIES OF CLAUSES}

\begin{definition}
  The body of a clause is defined by:
  \begin{itemize}
  \item $\arrayd{cstructure}{\maxNumNodes{}}{[0, \maxNumNodes{} - 1]}$ such
    that:
    \begin{itemize}
    \item $\variable{structure}[i] = i$: the $i$-th node is a root.
    \item $\variable{structure}[i] = j$: the $i$-th node's parent is node $j$.
    \end{itemize}
  \item $\arrayt{values}{\maxNumNodes{}}{Node}$.
  \end{itemize}
\end{definition}

\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (or) at (-0.375, 0) {$\lor$};
    \node[draw,circle,gray,text=black] (not) at (-1.5, -1) {$\neg$};
    \node[draw,circle,gray,text=black] (and) at (0.75, -1) {$\land$};
    \node[draw,gray,text=black] (P) at (-1.5, -2) {$\mathsf{P}(X)$};
    \node[draw,gray,text=black] (Q) at (0, -2) {$\mathsf{Q}(X)$};
    \node[draw,gray,text=black] (R) at (1.5, -2) {$\mathsf{P}(X)$};
    \draw[gray] (or) -- (not);
    \draw[gray] (or) -- (and);
    \draw[gray] (not) -- (P);
    \draw[gray] (and) -- (Q);
    \draw[gray] (and) -- (R);
  \end{tikzpicture}
  \caption{A tree representation of the formula from \cref{example:formula}}
  \label{fig:example_tree}
\end{figure}

\begin{example} \label{example:formula}
  Let $\maxNumNodes{} = 8$. Then $\neg\mathsf{P}(X) \lor (\mathsf{Q}(X)
  \land \mathsf{P}(X))$ corresponds to the tree in \cref{fig:example_tree} and
  can be encoded as:
  \begin{alignat*}{8}
    \variable{structure} &= [0, &&0, &&0, &&1, &&2, &&2, &&6, &&7],\\
    \variable{values} &= [{\lor}, &&{\neg}, &&{\land}, \mathsf{P}(&&X), \mathsf{Q}(&&X), \mathsf{P}(&&X), &&\top, &&\top],\\
    \variable{numNodes} &= 6, && && && && && && &&\\
    \variable{numTrees} &= 3. && && && && && && &&
  \end{alignat*}
  In the rest of this section, we will describe how the elements of
  $\variable{values}$ are encoded and list a series of constraints that make
  this representation unique.
\end{example}

\subsection{NODES}

@@

\subsection{CONSTRAINTS}

\begin{definition}
  Let $\variable{numTrees} \in \{ 1, \dots, \maxNumNodes{} \}$ be the number of
  trees in $\variable{structure}$, defined using the $\variable{tree}$
  constraint \citep{DBLP:conf/cp/FagesL11}.
\end{definition}

@@

\begin{constraint}
  $\variable{structure}[0] = 0$.
\end{constraint}

\begin{constraint}
  \variable{structure} is sorted.
\end{constraint}

\begin{constraint}
  For $i = 1, \dots, \maxNumNodes{} - 1$, if $i \ge \variable{numNodes}$, then
  \[
    \variable{structure}[i] = i, \quad \text{and} \quad
    \variable{values}[i].\variable{name} = \top,
  \]
  else $\variable{structure}[i] < i$.
\end{constraint}

\begin{constraint} \label{constraint:node_types}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let $C_i$ be the number of times $i$
  appears in the \variable{struture} array with index greater than $i$. Then
  \begin{align*}
    C = 0 &\iff \variable{values}[i].\variable{name} \in \predicates{} \cup \{ \top \},\\
    C = 1 &\iff \variable{values}[i].\variable{name} = \neg,\\
    C > 1 &\iff \variable{values}[i].\variable{name} \in \{ \land, \lor \},
  \end{align*}
  where $\variable{structure}_{>i}$ denotes array $\variable{structure}$ with positions
  greater than $i$.
\end{constraint}
Each of the three equivalences in \cref{constraint:node_types} corresponds to
node $i$ having no children, one child, and multiple children, respectively.

\begin{constraint}
  Only root nodes can have $\top$ as the value. For $i = 0, \dots,
  \maxNumNodes{} - 1$,
  \[
    \variable{structure}[i] \ne i \implies
    \variable{values}[i].\variable{name} \ne \top.
  \]
\end{constraint}

A way to `disable' a clause:
\begin{constraint}
  For $i = 0, \dots, \maxNumClauses{} - 1$, if
  $\variable{heads}[i].\variable{predicate} = \Box$, then
  \[
    \variable{bodies}[i].\variable{numNodes} = 1,
  \]
  and
  \[
    \variable{bodies}[i].\variable{values}[0].\variable{name} =
    \top.
  \]
\end{constraint}

\section{VARIABLE SYMMETRIES} \label{sec:variable_symmetry}

@@

  Then we can use the $\variable{setsIntsChanneling}$ constraint
  to define $\variable{occ}[|\constants{}| + |\variables{}| + 1]$ as an
  array of subsets of $\{ 0, \dots, N-1 \}$ such that for all $i = 0, \dots, N
  - 1$, and $t \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    i \in \variable{occ}[t] \quad \iff \quad
    \variable{terms}[i] = t
  \]
\end{definition}

We introduce an array that, for each variable, holds the position of its first
occurrence.
\begin{definition}
  Let $\arrayd{intros}{|\variables{}|}{\{ 0, \dots, N \}}$ be such that
  for $v \in \variables{}$,
  \[
    \variable{intros}[v] = \begin{cases}
      1 + \min \variable{occ}[v] & \text{if }
      \variable{occ}[v] \ne \emptyset\\
      0 & \text{otherwise.}
    \end{cases}
  \]
\end{definition}
Here, a value of zero means that the variable does not occur in the clause. We
want to use specifically zero for this so that we could use
\cref{constraint:diffbutzero} later. Because of this choice, the definition of
$\variable{intros}$ shifts all indices by one.

\begin{constraint}
  $\variable{intros}$ are sorted.
\end{constraint}

\begin{example}
  Let $\constants{} = \emptyset$, $\variables{} = \{ X, Y, Z \}$, $\maxArity{} =
  2$, $\maxNumNodes{} = 3$, and consider the clause
  \[
    \mathsf{sibling}(X, Y) \gets \mathsf{parent}(X, Z) \land
    \mathsf{parent}(Y, Z).
  \]
  Then $\variable{terms} = [X, Y, \Box, \Box, X, Z, Y, Z]$ (the boxes represent
  the conjunction node), $\variable{occ} = [\{ 0, 4 \}, \{ 1, 6 \},
  \{ 5, 7 \}, \{ 2, 3 \}]$, and $\variable{intros} = [0, 1, 5]$.
\end{example}

\subsection{REDUNDANT CONSTRAINTS}

@@

\begin{constraint}
  For $u \ne v \in \constants{} \cup \variables{} \cup \{ \Box \}$,
  \[
    \variable{occ}[u] \cap \variable{occ}[v] = \emptyset.
  \]
\end{constraint}

\begin{constraint} \label{constraint:diffbutzero}
  $\variable{allDifferentExcept0}(\variable{intros})$.
\end{constraint}

\begin{constraint}
  For $v \in \variables{}$,
  \[
    \variable{intros}[v] \ne 0 \quad \iff \quad
    \variable{intros}[v] - 1 \in \variable{occ}[v].
  \]
\end{constraint}

\begin{definition}
  We define an auxiliary set variable
  \[
    \variable{potentialIntros} \subseteq \{ 0, \dots, N \}
  \]
  to act as a set of possible values that $\variable{intros}$ can take,
  i.e., for $v \in \variables{}$, $\variable{intros}[v] \in
  \variable{potentialIntros}$.
\end{definition}

\begin{constraint} \label{constraint:potentialIntroductions}
  For $i = 0, \dots, \maxNumNodes{} - 1$, let
  \[
    S = \{ \maxArity{} \times (i + 1) + j + 1 \mid j = 0, \dots, \maxArity{} - 1.
    \}
  \]
  If $\variable{values}[i].\variable{name} \not\in \predicates{}$, then
  \[
    \variable{potentialIntros} \cap S = \emptyset.
  \]
\end{constraint}
In simpler terms, \cref{constraint:potentialIntroductions} says that if a node
in the tree representation of a clause represents something other than a
predicate, then a variable cannot be introduced as one of its `arguments'.

\section{COUNTING PROGRAMS}

@@

We will first consider clauses with gaps, i.e., without taking variables and
constants into account. Let $T(n, a)$ denote the number of possible clause
bodies with $n$ nodes and total arity $a$. Then $T(1, a)$ is the number of
predicates in $\predicates{}$ with arity $a$, and the following recursive
definition can be applied for $n > 1$:
\begin{align*}
  T(n, a) = T(n-1, a) + 2&\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
  c_i \ge 1 \text{ for all } i}}\\
  &\sum_{\substack{d_1 + \dots + d_k = a,\\
  d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\end{align*}
The first term here represents negation, i.e., negating an expression consumes
one node but otherwise leaves the task unchanged. If the first operation is not
negation, then it must be either conjunction or disjunction (hence the
coefficient `2'). In the first sum, $k$ represents the number of children of the
root node, and each $c_i$ is the number of nodes dedicated to child $i$. Thus,
the first sum iterates over all possible ways to partition the remaining $n-1$
nodes. Similarly, the second sum considers every possible way to partition the
total arity $a$ across the $k$ children nodes.

@@

The number of ways to fill $n$ gaps with terms can be expressed as
\begin{align*}
  P(n) = |\constants{}|^n + &\sum_{\substack{1 \le k \le |\variables{}|, \\ 0 =
      s_0 < s_1 < \dots < s_k < s_{k+1} = n+1}}\\
  &\prod_{i=0}^k (|\constants{}| + i)^{s_{i+1} - s_i - 1}.
\end{align*}
The first term is simply the number of ways to fill all $n$ gaps with
constants. The parameter $k$ is the number of variables used in the clause, and
$s_1, \dots, s_k$ mark the first occurrence of each variable. For each gap
between any two introductions (or before the first introduction, or after the last
introduction), we have $s_{i+1}-s_i-1$ spaces to be filled with any of the
$\constants{}$ constants or any of the $i$ already-introduced variables.

@@

\section{PREDICATE INDEPENDENCE}

@@

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
    \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
    \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
    \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
    \draw[gray] (and) -- (parent);
    \draw[gray] (and) -- (not);
    \draw[gray] (not) -- (mother);
  \end{tikzpicture}
  \caption{The tree representation of the body of Clause~\eqref{eq:example_clause}.}
  \label{fig:example_tree2}
\end{figure}
\begin{figure}
  \centering
  \begin{tikzpicture}
    \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
    \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
    \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
    \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
    \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
    \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
  \end{tikzpicture}
  \caption{The predicate dependency graph of the program in \cref{ex:program}.
    Positive edges are labeled with `$+$', and negative edges with `$-$'.}
  \label{fig:predicate_dependencies}
\end{figure}

@@

  \For{$(d_1, d_2) \in \getDependencies{$p_1$, 0} \times \getDependencies{$p_2$,
      0}$ such that $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
